Active Learning from Blackbox to Timed Connectors

### 2016-01-20 paper/paper.tex
.

% finally no todos should exist in this draft
\usepackage[textsize=tiny]{todonotes}

@@

\begin{document}
\maketitle 
\begin{abstract}
  Coordination models and languages play a key role in formally specifying the communication and
  interaction among different components in large-scale distributed and concurrent systems. In this
  paper, we propose an active learning framework to extract timed connector models from black-box
  system implementation. 
  We first introduce parameterized Mealy machine(PMM) as an operational semantic
  model for channel-based coordination language Reo. PMM serves as a bridge
  between Reo connectors and Mealy machines. With the product operator, complex connectors can be
  constructed by joining basic channels and the PMM of connectors can be transformed into Mealy
  machines. Moreover, we adapt L*, a well-known learning algorithm, to timed connectors (in the form
  of Mealy machines). The new algorithm has shown its efficiency in multiple case studies. 
  This framework has been implemented as a package in \texttt{Golang}.
\end{abstract}

@@

\emph{Correctness} of connectors is closely related to some low-level implementation details.
For example, well-written code may behave dramatically weird with an improper set of concurrency
primitives. Such scenarios happen frequently in embedded systems with different hardware platforms or
operating systems. Consequently, manually extracting a proper model for an existing connector
implementation seems rather unreliable, even with reference to its source code.

@@

Various forms of Mealy machines have been defined in the literature. 
Since active automata learning requires the system-under-learn to be deterministic, in
this paper we take the deterministic Mealy machine model as defined in 
\cite{DBLP:conf/sfm/SteffenHM11}.

\begin{definition}[Mealy machine]
  A Mealy machine is a 6-tuple $(S, s_0, I, O, \delta, \lambda)$ consisting of:
  \begin{itemize}
    \item[-] a finite set of states $S$
    \item[-] an initial $s_0\in S$
    \item[-] a finite set of inputs $I$
    \item[-] a finite set of outputs $O$
    \item[-] an output function $\delta : S \times  I \rightarrow O$ mapping a pair
      of a state and an input to the corresponding output.
    \item[-] a transition function $\lambda : S \times I \rightarrow S$ mapping a pair of a
      state and an input to the corresponding successor state.
  \end{itemize}
\end{definition}

@@

\section{Timed Connectors as Mealy Machines}
\label{sec:semantics}
In this section, we illustrate how to transform a timed connector into a Mealy machine with timed
action.
Since time is not involved in original Mealy machine, we first discuss how to formalize the time
dimension in Mealy machine. After that, we present the notion of \emph{parameterized Mealy
machine} as a bridge between connectors and Mealy machines.

@@

With \emph{link} and \emph{prop} defined, a connector can be formed by simpler ones in
form of \emph{parameterized Mealy machines}. It will be transformed into a concrete Mealy machine
later, once the alphabet $\Sigma$ is provided.

@@

\subsection{Observation Table}
\emph{Observation Tables}, proposed in \cite{DBLP:journals/iandc/Angluin87}, is used to construct
a hypothetical Mealy machine from the results of membership queries.

Generally, a Mealy machines is determined by its \emph{states} and \emph{transitions}.
As for blackboxes, input sequence $s\in\mathcal{I}^*$ can be used to denote its state 
after execution of $s$. Similarly, for all $a\in\mathcal{I}$, $sa$ indicates the successor
state of $s$ after execution of $a$.
However, under such notations, there are \emph{infinite} number of states where most are
reduplicate.

States are distinguished, usually, by their \emph{subsequent behavior}. Suppose $s_1\neq s_2\in
\mathcal{I}^*$, the corresponding of $s_1$ and $s_2$ are exactly the same iff $\forall d\in
\mathcal{I}^+. mq(s_1d)=mq(s_2d)$. Since we can never check it on every $d\in\mathcal{I}^+$, an
alternative way is to use a set of \emph{suffixes} to distinguish them approximately.
\begin{definition}[$D$-Equivalence]
  provided with two access sequences $s_1,s_2$ and a set of suffixes $D\subset\mathcal{I}^+$, we say
  the corresponding states of $s_1$ and $s_2$ are $D$-equivalent (denoted as $s_1\sim_D s_2$)
  iff $\forall d\in D, mq(s_1d) = mq(s_2d)$.
\end{definition}
\begin{definition}[Access Sequence]
  Given an input sequence $s$, we define the access sequence of $s$ as
  \[
  acc(s,D)=\min\{s'\in\mathcal{I}^*|s'\sim_D s\}
  \]
\end{definition}

Formally, an observation table is defined as a tuple $(S,D)$ where $S\subset\mathcal{I}^*$ indicates
the access sequences of states and $D\subset\mathcal{I}^*$ indicates the suffixes. Its
tabular form can be found in Figure \ref{fig:hypo}. In such tables, columns are labelled with
$d\in D$, and rows are labelled with input sequences. Rows are divided into two parts. The upper
part, labelled with $s\in S$, shows the states in current hypothesis (in form of access sequences).
The lower part, labelled with $sa$ (where $s\in S,a\in\mathcal{I}$) indicates all the transitions
from $S$. A cell with row label $s$ and column label $d$ is filled with $mq(sd)$.

@@

    \emph{(we use input sequences to indicate their corresponding states in following parts)}
    \todo{define the empty sequence}
  \item[-] $\delta(s,a)=mq(sa)$
  \item[-] $\lambda(s,a)=acc(sa,obs.D)$
\end{itemize}

@@

\begin{algorithm} 
  \caption{CloseTable} 
  \label{alg:buildtable}
  \small
  \KwIn{Oracle interface $mq$, Input actions $\mathcal{I}$, suffix set $D$}
  \KwOut{Observation table $obs$} 
  $obs$ initialized as empty\;
  $unclosed=\{\varepsilon\}$\;
  \Repeat{$unclosed=\varnothing$}
  {
    $next=\{st|\forall s\in unclosed,\forall t\in \mathcal{I}\}$\;
    append $unclosed$ to $obs$\;
    $unclosed=\{seq\in next|\forall a\in obs, seq \not\sim_D a\}$\;
  }
  \Return $obs$\; 
\end{algorithm}

@@

Firstly, $obs$ is initialized as empty and the empty sequence $\varepsilon$ is pushed in $unclosed$.
Then we will explore all the access sequences in $unclosed$ and calculate its successors. Here the
$unclosed$ set consists of access sequences that has no equivalent fellow in $S$. 
For example, figure \ref{fig:hypo} shows the $obs$ of a 2-Timer channel after the first
iteration. The five successors of $\varepsilon$ is presented as the bottom part of the table, where
four are equivalent with $\varepsilon$ but $a,\rnoread$ is not. Therefore, we take $a,\rnoread$ as a
brand-new state and all of its successors need further exploration (see in the hypothesis Mealy
machine presented in the following figure).

@@

In the second iteration, we explore the successors of $a,\rnoread$. Fortunately we find that every
successor has a shorted equivalence fellow (\emph{itself}). Consequently, the algorithm terminates
with a closed hypothesis where all the unclosed edge shown in Figure \ref{fig:hypo} turn into
self-loop.

@@

The whole L* algorithm can be concluded as follows.
\begin{algorithm} 
  \caption{L*} 
  \label{alg:lstar}
  \KwIn{Oracle interface $mq,eq$, Input actions $\mathcal{I}$}
  \KwOut{Observation table $obs$} 
  $D=I$\;
  \Repeat{$ce=true$}{
    $obs$ = CloseTable($mq$, $\mathcal{I}$, $D$)\;
    $ce=eq(obs)$\;
    \If {$ce!=true$} {
      D.append(MaxSuffix(s))\;
    }
  }
  \Return $obs$\; 
\end{algorithm}

@@

\section{Conclusion and Future Work}
In this paper, we come up with an approach to learn timed connectors from blackboxes. We define
the \emph{parameterized Mealy machine} to serve as an operational semantics of Reo. Then we
describe a set of basic channels with parameterized Mealy machine, which can be used to construct
complex connectors by means of the production operator and link operator.
Besides, we show that every connector can be transformed into a Mealy
machine with time action T, and such Mealy machines can be extracted by an improved version of
well-knwon active learning algorithm L*.

The original L* algorithm run into the bottleneck when time is involved. By a tree-style cache, we
make significant reduction on membership queries and, in turn, improve the performance of learning
algorithm. As a by-product, we also encapsulate the Reo connector as a distributable package which
could contribute to concurrent programming.

>>>>

% finally no todos should exist in this draft
% \usepackage[textsize=tiny]{todonotes}

@@

\begin{document}
\maketitle 
\begin{abstract}
  Coordination models and languages play a key role in formally specifying the communication and
  interaction among different components in large-scale distributed and concurrent systems. In this
  paper, we propose an active learning framework to extract timed connector models from black-box
  system implementation. 
  We first introduce parameterized Mealy machine(PMM) as an operational semantic
  model for channel-based coordination language Reo. PMM serves as a bridge
  between Reo connectors and Mealy machines. With the product operator, complex connectors can be
  constructed by joining basic channels and the PMM of connectors can be transformed into Mealy
  machines. Moreover, we adapt L*, a well-known learning algorithm, to timed connectors (in the form
  of Mealy machines). The new algorithm has shown its efficiency in multiple case studies. 
  This framework has been implemented in \texttt{Golang}.
\end{abstract}

@@

\emph{Correctness} of connectors is closely related to some low-level implementation details.
For example, well-written code may behave dramatically weird with an improper set of concurrency
primitives. Such scenarios happen frequently in embedded systems with different hardware platforms or
operating systems. Consequently, manually extracting a proper model from an existing connector
implementation seems rather unreliable, even with reference to its source code.

@@

Various forms of Mealy machines have been defined in the
literature\cite{George1955A,DBLP:journals/jcss/Broy14,DBLP:conf/sfm/SteffenHM11}. 
Since active automata learning requires the system-under-learn to be deterministic, in
this paper we take the deterministic Mealy machine model as defined in 
\cite{DBLP:conf/sfm/SteffenHM11}.

\begin{definition}[Mealy machine]
  A Mealy machine is a 6-tuple $(S, s_0, I, O, \delta, \lambda)$ consisting of:
  \begin{itemize}
    \item[-] a finite set of states $S$
    \item[-] an initial state $s_0\in S$
    \item[-] a finite set of inputs $I$
    \item[-] a finite set of outputs $O$
    \item[-] an output function $\delta : S \times  I \rightarrow O$ mapping a pair
      of a state and an input to the corresponding output.
    \item[-] a transition function $\lambda : S \times I \rightarrow S$ mapping a pair of a
      state and an input to the corresponding successor state.
  \end{itemize}
\end{definition}

@@

\section{Timed Connectors as Mealy Machines}
\label{sec:semantics}

In this section, we illustrate how to transform a timed connector into a Mealy machine with timed
action.
Since time is not involved in original Mealy machine, we first discuss how to formalize the time
dimension in Mealy machine. After that, we present the notion of \emph{parameterized Mealy
machine} as a bridge between connectors and Mealy machines.

@@

With \emph{link} and \emph{prop} defined, connectors can be constructed by simpler ones in
form of PMMs. It will be transformed into a concrete Mealy machine later, once the alphabet $\Sigma$
is provided.

@@

\subsection{Observation Table}
\emph{Observation Tables}, proposed in \cite{DBLP:journals/iandc/Angluin87}, are used to construct
hypothetical Mealy machines from the results of membership queries.

Mealy machines are composed of \emph{states} and \emph{transitions}.
As for blackboxes, input sequence $s\in\mathcal{I}^*$ can be used to denote its state 
after executing $s$. Similarly, for all $a\in\mathcal{I}$, we use $sa$ to denote the successor
state of $s$ after executing $a$.
However, under such notations, there are \emph{infinite} number of states, among which most are
reduplicate.

States are distinguished, usually, by their \emph{subsequent behavior}. Suppose $s_1\neq s_2\in
\mathcal{I}^*$, we say the corresponding of states $s_1$ and $s_2$ are equivalent iff $\forall d\in
\mathcal{I}^+. mq(s_1d)=mq(s_2d)$. Since we can never check it on every $d\in\mathcal{I}^+$, an
alternative way is to use a set of \emph{suffixes} to distinguish them approximately.
\begin{definition}[$D$-Equivalence]
  provided with a set of suffixes $D\subset\mathcal{I}^+$, we say the corresponding states of
  $s_1,s_2\in\mathcal{I}^*$ are $D$-equivalent (denoted as $s_1\sim_D s_2$)
  iff $mq(s_1d) = mq(s_2d)$ for all $d\in D$.
\end{definition}

Now we use \emph{access sequence} to describe the states. Given an input sequence $s$, the access
sequence of $s$ is defined as
\[
acc(s,D)=\min\{s'\in\mathcal{I}^*|s'\sim_D s\}
\]

Formally, an observation table (see in Figure \ref{fig:hypo}) is determined by a tuple $(S,D)$ where
$S\subset\mathcal{I}^*$ is a set of the access sequences of states and $D\subset\mathcal{I}^*$
is a set of suffixes. In such tables, columns are labelled with suffixes in $D$, and rows are
labelled with input sequences. As shown in the example, rows are divided into two parts. The upper
part, labelled with access sequence $s\in S$, denotes the states covered in the current hypothesis.
The lower part, labelled with sequences in $\{sa|s\in S,a\in\mathcal{I}\}\backslash S$ (representing
any potentially unclosed transition) denotes all the transitions outside $s\in S$. A cell with row
label $s$ and column label $d$ is filled with $mq(sd)$.

@@

    \emph{(we use input sequences to indicate their corresponding states in following parts)}
  \item[-] $\delta(s,a)=mq(sa)$
  \item[-] $\lambda(s,a)=acc(sa,obs.D)$. Since $obs$ is closed, we always have $\lambda(s,a)\in S$.
\end{itemize}

@@

\begin{algorithm} 
  \caption{CloseTable} 
  \label{alg:buildtable}
  \small
  \KwIn{An oracle interface $mq$, An input actions $\mathcal{I}$, An observation table $obs$}
  \KwOut{A closed observation table $obs$} 
  \Repeat{$unclosed=\varnothing$}
  {
    $next=\{st|\forall s\in obs.S,\forall t\in \mathcal{I}\}\backslash obs.S$\;
    $unclosed=\{seq\in next|\forall a\in obs.S, seq \not\sim_{obs.D} a\}$\;
    $obs.S$ = $obs.S\cup unclosed$\;
  }
  \Return $obs$\; 
\end{algorithm}

@@

Firstly, $obs.S$ is initialized with the initial state (denoted by $\varepsilon$), and $obs.D$ is
initialized with all one-step suffixes.
Then we will explore all the access sequences in $obs.S$ and calculate its successors. Here the
$unclosed$ set consists of access sequences that has no equivalent fellow in $obs.S$. 
For example, figure \ref{fig:hypo} shows the $obs$ of a 2-Timer channel after the first
iteration. The five successors of $\varepsilon$ is presented as the bottom part of the table, where
four are equivalent with $\varepsilon$ but $a,\rnoread$ is not. Therefore, we take $a,\rnoread$ as a
brand-new state and all of its successors need further exploration (see in the hypothesis Mealy
machine presented in the following figure).

@@

In the second iteration, we explore the successors of $a,\rnoread$. Fortunately, now every
successor has a equivalence fellow in $obs.S$ (\emph{itself}). Consequently, the algorithm terminates
with a closed hypothesis where all the unclosed edge shown in Figure \ref{fig:hypo} turn into
self-loop.

@@

The whole L* algorithm can be concluded as follows.
\begin{algorithm} 
  \caption{L*} 
  \label{alg:lstar}
  \KwIn{Oracle interface $mq,eq$, Input actions $\mathcal{I}$}
  \KwOut{Observation table $obs$} 
  $obs.S$ initialized as empty\;
  $obs.D$ initialized as $\{\{i\}|i\in\mathcal{I}\}$\;
  \Repeat{$ce=true$}{
    $obs$ = CloseTable($mq$, $\mathcal{I}$, $D$)\;
    $ce=eq(obs)$\;
    \If {$ce!=true$} {
      D.append(MaxSuffix(s))\;
    }
  }
  \Return $obs$\; 
\end{algorithm}

@@

\section{Conclusion and Future Work}
In this paper, we come up with an approach to extract timed connectors from blackbox
implementations. We propose \emph{parameterized Mealy machine} to describe the behavior of Reo
connectors. Then we define the product operator and link operator, which can be used to
construct complex connectors from simpler ones.
Besides, we show how PMMs behave as the bridge between Reo connectors and concrete Mealy
machines with timed action T provided.

We also adapt the well-knwon active learning algorithm L* to deal with time domain.
When time is taken into consideration, the original L* algorithm run into the bottleneck.
By a tree-style cache, we make significant reduction on membership queries and, in turn,
improve the performance of learning algorithm. As a by-product, we also encapsulate the Reo
connector as a distributable package which could contribute to concurrent programming.
